---
title: "Impact of long-COVID on NBA point productivity"
author: "Kent"
date: "6/18/2021"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
library(knitr)
library(tidyverse)
library(dplyr)
library(rvest)
library(igraph)
library(lfe)
library(kableExtra)
library(nbastatR)
library(dummies)
library(MatchIt)
library(quantreg)
library(here)

# clean environment
rm(list = ls())
```

```{r clean the injury dataframe, message=FALSE}
## Clean the injury dataframe

NBACovidTable <- read.csv(here("data/NBACovidTable.csv"), 
                          na.strings = c("NA",""," "))

NBACovidTable <- filter(NBACovidTable, !is.na(Relinquished))


### Get rid of non-alphanumeric symbols
playerNames <- gsub("• ", "", NBACovidTable$Relinquished)

NBACovidTable$Relinquished <- playerNames
rm(playerNames)


# Filter for notes containing COVID-19
NBACovidTable <- NBACovidTable[grepl("COVID-19", NBACovidTable$Notes) |
                                 grepl("protocols", NBACovidTable$Notes),]

# Convert to date form
NBACovidTable$Date <- as.Date(NBACovidTable$Date)

# Just get the last date
NBACovidTable <- NBACovidTable %>% 
  group_by(Relinquished) %>%
  arrange(Date) %>% 
  mutate(protocol = 1) %>% 
  mutate(numProtocol = cumsum(protocol)) %>% 
  mutate(lastProtocol = ifelse(numProtocol == max(numProtocol), 1, 0)) %>% 
  filter(lastProtocol == 1)

# Get rid of the nick names
## "\\" escapes for special characters
## "." is all characters
## "*" is zero or more
playerNames <- gsub(" \\/.*","", NBACovidTable$Relinquished)
NBACovidTable$Relinquished <- playerNames
rm(playerNames)

```


```{r player stats, message=FALSE}


## Get game ids
NBAgames <- game_logs(seasons = 2021)

NBAgames <- select(NBAgames, c("dateGame","nameTeam","locationGame","slugTeam","slugOpponent",
                                "outcomeGame","namePlayer","minutes",
                               "treb","ast","stl","blk","tov","pf","pts","plusminus"))


# Merge the dataframes together

df1 <- merge(NBAgames, NBACovidTable, by.x = "namePlayer", by.y = "Relinquished",
             all.x = T, all.y = F)

# Keeping just df1
rm(list=setdiff(ls(), "df1"))

# Create some variables
# Get time since event variable
# Get time since event variable
df1 <- df1 %>% 
  group_by(slugTeam) %>% 
  mutate(Dk = ifelse(is.na(Date), NA, as.Date(dateGame) - as.Date(Date)),
         COVID = ifelse(is.na(Date), 0, 1)) %>% 
  arrange(Dk)


# Creating the dummies
dayDummies <- dummy(df1$Dk)


# Joining the matrix to the dataframe
dfEvent <- data.frame(df1, dayDummies)



```



### Introduction

There is some uncertainty on the effects of long-COVID on work productivity (CITE). We estimate the impact of long-COVID on trends in productivity through changes in points per game in the NBA. At the highest level of professional sports, athletes may be one of the closest things we can observe to the limit of physical performance. Competing at this level can also be demanding mentally. If long-COVID effects productivity through physical or mental wellbeing then we assume this would be observable through the performance of top-level athletes in the NBA.

Unlike many other sports, players often score many points during a game in the NBAm giving us more variability. Additionally, the NBA plays 72 games in a season and basketball in particular requires a substantial sample size and increases the likelihood of observing any effects.

Although we have the we have the benefit of having an abundance of data on NBA statistics available, one of our main limitations is that the NBA releases the names of the players who have had COVID-19. Instead we use injury reports showing when players were placed on covid protocol. This shows us the players who have potentially been exposed to the virus, but not neccessarily those who have had it. Therefore, we treat our estimates as a lower bound.

### Descriptive Statistics

From Table 1, we observe that players who have been placed on covid protocol typically score
fewer points.

```{r make table}


# Convert characters to factors
names <- c("namePlayer", "dateGame", 
           "locationGame", "slugTeam", 
           "slugOpponent")
dfEvent[,names] <- lapply(dfEvent[,names], factor) # lapply functions like a for loop
rm(names)

# Descriptive stuff
dfEvent$`COVID protocol` <- ifelse(dfEvent$COVID == 1, "yes", "no")

tbl1 <- dfEvent %>% 
  group_by(`COVID protocol`) %>% 
  summarise(players = n(),
            `mean pts` = mean(pts),
            `std error` = sd(pts)/sqrt(players))

kable(tbl1, "html", digits = 2, caption = "General Summary Statistics", booktabs = T, linesep = "") %>% 
  kable_styling(latex_options = "hold_position")

```

#### Raw data trends

Using a weight polynomial regressions (LOESS) with a span of 5% of the data we show the trends
in points from players who were placed on covid protocol. From these results, there does not
appear to be a trend in any direction after the players were placed on protocol.

```{r raw data trends}
df1 %>% filter(!is.na(Date)) %>% 
  ggplot(mapping = aes(x = Dk, y = pts)) +
  geom_point(colour = "snow3", size = 0.1) +
  geom_smooth(aes(x = Dk, y = pts),
              span = 0.10,
              se = T,
              fill = "skyblue",
              method = "loess") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "lightgrey") +
  geom_vline(aes(xintercept = 0,
                 colour = "indianred4"),
             size = 0.1) +
  guides(colour = F) +
  labs(title = "Points in relation to COVID-19 protocol",
       x = "days from COVID-19 protocol",
       y = "points") +
  theme_bw()
```

### Event study

To estimate the impact of long-covid on points per game for NBA players we use the following model:

(write model in latex)

where:

(more latex code)

### Results

From Figure 2, it doesn’t look like there is any trend at all. Maybe this is because long covid
only effect around 1.5% of people and we observe 231 players who were placed on protocol. Who
knows...

```{r event study bins}

# Create bins
dfEvent <- dfEvent %>%
  mutate(lower_bin = select(dfEvent, Dk.147:Dk.11) %>%
           rowSums(na.rm = TRUE))    # 10 days of lags

dfEvent <- dfEvent %>%
  mutate(upper_bin = select(dfEvent, Dk51:Dk144) %>%
           rowSums(na.rm = TRUE))    # 50 days of leads
```

```{r event study basic}
# Run regression
reg1 <- felm(pts ~
                    # Time relative to events
                    lower_bin +
                    Dk.10 + Dk.9 + Dk.8 + Dk.7 + Dk.6 + Dk.5 + Dk.4 + Dk.3 + Dk.2 + Dk.1 +
                    Dk1 + Dk2 + Dk3 + Dk4 + Dk5 + Dk6 + Dk7 + Dk8 + Dk9 + Dk10 +
                    Dk11 + Dk12 + Dk13 + Dk14 + Dk15 + Dk16 + Dk17 + Dk18 + Dk19 +
                    Dk20 + Dk21 + Dk22 + Dk23 + Dk24 + Dk25 + Dk26 + Dk27 + Dk28 +
                    Dk29 + Dk30 + Dk31 + Dk32 + Dk33 + Dk34 + Dk35 + Dk36 + Dk37 +
                    Dk38 + Dk39 + Dk40 + Dk41 + Dk42 + Dk43 + Dk44 + Dk45 + Dk46 +
                    Dk47 + Dk48 + Dk49 + Dk50 +
                    upper_bin
                  |
                    # Facility, year and day-of-year fixed-effects
                    dateGame + locationGame + slugTeam + slugOpponent  |0|
                    # No clustering
                    0,
                  weights = NULL,
                  data = dfEvent)


# Creating horizontal variable for figure
x <- c(-10:-1, 1:50)


# Getting the coefficients
coef1 <- coef(reg1)[2:61]

# Collecting the standard errors
se <- as.numeric(summary(reg1, robust = F)$coef[2:61,2])


# Normalizing origin
origin <- data.frame(x = 0, coef1 = 0, se = 0)


# Getting the event for plot
event1 <- data.frame(x = x, coef1 = coef1, se = se)
event1 <- rbind(event1, origin)


ggplot(event1, aes(x = x, y = coef1)) +
  geom_linerange(aes(ymin = coef1 - 1.96*se, ymax = coef1 + 1.96*se),
                color = "royalblue") +
  geom_point(color = "royalblue") +
  geom_vline(xintercept = 0, color = "firebrick3") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  labs(x = "days from protocol",
       y = "change in points",
       title = "Effect of COVID-19 on Player Point Productivity") +
  # scale_x_continuous(breaks = seq(-10,100,10)) +
  theme_bw()


```


#### Quantile regression
Maybe there isn’t enough variation from the players with low point productivity. Lets try a
quantile regression using a similar model to Model 1 using τ = 0.75. I don’t know about these
results either.

```{r quantile regression}
# # Try quantile regression
# Run regression
# quant1 <- rq(pts ~
#                # Time relative to events
#                lower_bin +
#                Dk.10 + Dk.9 + Dk.8 + Dk.7 + Dk.6 + Dk.5 + Dk.4 + Dk.3 + Dk.2 + Dk.1 +
#                Dk1 + Dk2 + Dk3 + Dk4 + Dk5 + Dk6 + Dk7 + Dk8 + Dk9 + Dk10 +
#                Dk11 + Dk12 + Dk13 + Dk14 + Dk15 + Dk16 + Dk17 + Dk18 + Dk19 +
#                Dk20 + Dk21 + Dk22 + Dk23 + Dk24 + Dk25 + Dk26 + Dk27 + Dk28 +
#                Dk29 + Dk30 + Dk31 + Dk32 + Dk33 + Dk34 + Dk35 + Dk36 + Dk37 +
#                Dk38 + Dk39 + Dk40 + Dk41 + Dk42 + Dk43 + Dk44 + Dk45 + Dk46 +
#                Dk47 + Dk48 + Dk49 + Dk50 +
#                upper_bin +
#                # Facility, year and day-of-year effects
#                dateGame + locationGame + slugTeam + slugOpponent,
#              weights = NULL,
#              tau = 0.75,
#              data = dfEvent)
# 
# # Creating horizontal variable for figure
# x <- c(-10:-1, 1:50)
# 
# 
# # Getting the coefficients
# coef1 <- coef(quant1)[2:61]
# 
# # Collecting the standard errors
# se <- as.numeric(summary(quant1, robust = F)$coef[2:61,2])
# 
# 
# # Normalizing origin
# origin <- data.frame(x = 0, coef1 = 0, se = 0)
# 
# 
# # Getting the event for plot
# event1 <- data.frame(x = x, coef1 = coef1, se = se)
# event1 <- rbind(event1, origin)
# 
# 
# ggplot(event1, aes(x = x, y = coef1)) +
#   geom_linerange(aes(ymin = coef1 - 1.96*se, ymax = coef1 + 1.96*se),
#                  color = "royalblue") +
#   geom_point(color = "royalblue") +
#   geom_vline(xintercept = 0, color = "firebrick3") +
#   geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
#   labs(x = "days from protocol",
#        y = "change in points",
#        title = "Effect of COVID-19 on Player Point Productivity",
#        subtitle = "For Players in the 75th Percentile of Scoring") +
#   # scale_x_continuous(breaks = seq(-10,100,10)) +
#   theme_bw()


```


